---
title: "extended_exercise"
format: html
editor: visual
---

## Connecting to the Yahoo finance API

The NOK/EUR exchange rate data is sourced from Yahoo Finance using the *quantmod* package in R. The dataset spans from January 1, 2015, to January 1, 2025, and includes the following variables: Date, Open, High, Low, Close, Volume, and Adjusted Close.

```{r}
library(quantmod)
library(ggplot2)
library(tseries)
library(dplyr)

# We retrieve NOK/EUR daily exchange rate data from 2015 to 2025 using Yahoo Finance API
getSymbols("NOKEUR=X", src = "yahoo", from = "2015-01-01", to = "2025-01-01") 

# We convert the data to dataframe and adjust column names
nokeur_data <- data.frame(Date = index(`NOKEUR=X`), coredata(`NOKEUR=X`))
nokeur_data <- na.omit(nokeur_data)
colnames(nokeur_data) <- c("Date", "Open", "High", "Low", "Close", "Volume", "Adjusted")

ggplot(nokeur_data, aes(x = Date, y = Close)) +
  geom_line(color = "darkorchid4") + 
  labs(title = "NOK/EUR Exchange Rate", 
       x = "Date", y = "Exchange Rate") +
  theme_minimal()
```

Since many time-series models require stationarity, log returns of the exchange rate are computed using:

$$
LogReturn_t = \log \left( \frac{Close_t}{Close_{t-1}} \right).
$$

This transformation helps in stabilizing variance and making the series more stationary.

```{r}
# Most time-series models require stationary data. Log returns help achieve stationarity.
nokeur_data <- nokeur_data %>%
  mutate(Log_Returns = log(Close / lag(Close)))

# Remove NA values caused by lag
nokeur_data <- na.omit(nokeur_data)

# We plot the data to visualy inspect stationarity, volatility clustering,...
ggplot(nokeur_data, aes(x = Date, y = Log_Returns)) +
  geom_line(color = "cyan4") +
  labs(title = "Log Returns of NOK/EUR Exchange Rate", 
       x = "Date", y = "Log Returns") +
  theme_minimal()

```

A time-series plot of the NOK/EUR exchange rate shows trends and fluctuations over time. However, when log returns are plotted, they appear more stationary, with visible volatility clustering---periods of high and low fluctuations in returns.

```{r}
# We confirm stationarity with ADF test (H0 = Time series is non-stationary)
adf.test(nokeur_data$Log_Returns, alternative = "stationary")
```

To formally test stationarity, the **Augmented Dickey-Fuller (ADF) test** is conducted. The null hypothesis (H₀) assumes that time series is non-stationary, while the alternative hypothesis (H₁) assumes stationarity. Since the p-value of the ADF test is less than 0.05, we reject H₀ and conclude that log returns are stationary, making them suitable for time-series modeling.

```{r}
# ACF and PACF of Log Returns
par(mfrow = c(1, 2))
acf(nokeur_data$Log_Returns, main = "ACF of Log Returns")
pacf(nokeur_data$Log_Returns, main = "PACF of Log Returns")
```

Autocorrelation (ACF) and Partial Autocorrelation (PACF) plots of log returns are examined to detect linear dependencies. The results indicate minimal autocorrelation, implying that past returns do not significantly influence future returns. Consequently, simple autoregressive models such as AR(1) are not useful in this case.

```{r}
# ACF and PACF of squared Log Returns to check for volatility clustering and non-linear dependence
par(mfrow = c(1, 2))
acf(nokeur_data$Log_Returns^2, main = "ACF of Squared Log Returns")
pacf(nokeur_data$Log_Returns^2, main = "PACF of Squared Log Returns")

```

To analyze non-linear dependencies, ACF and PACF plots of **squared log returns** are examined. The presence of significant autocorrelation suggests volatility clustering---periods of high volatility are followed by high volatility, and low volatility by low volatility.

```{r}
# squared returns show significant autocorrelation -> we confirm that with the following test:
lags <- 1:10
ljung_box_results <- data.frame(
  lag = lags,
  Q_statistic = sapply(lags, function(lag) Box.test(
    nokeur_data$Log_Returns^2,
    lag = lag,
    type = "Ljung-Box")$statistic),
  p_value = sapply(lags, function(lag) Box.test(
    nokeur_data$Log_Returns^2,
    lag = lag,
    type = "Ljung-Box")$p.value)
)
ljung_box_results
```

To statistically confirm volatility clustering, the **Ljung-Box test** is performed on squared log returns. The null hypothesis (H₀) assumes no autocorrelation, meaning no volatility clustering. Since all p-values are close to zero, we reject H₀, confirming that volatility clusters over time. Therefore, models designed for capturing volatility, such as **GARCH, TARCH, or FIGARCH**, are more appropriate for forecasting short-term fluctuations. 
